\documentclass{article}
\usepackage{graphicx} % Required for inserting images
\usepackage{titlesec}
\usepackage[compress]{natbib}
\usepackage[resetlabels]{multibib}
\usepackage{authblk}
\usepackage{amsmath,amssymb,bm}
\usepackage{braket}

\usepackage[table]{xcolor}
\usepackage{url}            % simple URL typesetting
\usepackage{float}
\definecolor{dred}{rgb}{0.6,0,0}
\definecolor{dpurple}{HTML}{A020F0}
\definecolor{dblue}{rgb}{0,0,0.6}
\definecolor{hlcolor}{rgb}{1,1,0.8}
\usepackage[colorlinks=true, linkcolor=dblue, citecolor=dred, urlcolor=dblue]{hyperref}       % hyperlinks

\usepackage[nameinlink]{cleveref}
\usepackage[margin=1in]{geometry}
\renewcommand\b\bm
\newcommand*\sgn[1]{\text{sgn} ({#1})}
\setlength{\parindent}{0pt}
\setlength{\parskip  }{5.5pt}
\creflabelformat{equation}{#2\textup{#1}#3}

\title{\Large{Learning to predict model-based actions}}

\author[1]{\normalsize \bfseries Kristopher T.\ Jensen}
\affil[1]{ \small Sainsbury Wellcome Centre, University College London, London, UK}
\date{\vspace*{-2.0em}
\normalsize kris.torp.jensen@gmail.com
}



\begin{document}

\maketitle

\subsection*{Task}

We consider the simple task of predicting the binary action of a linear teacher network from a $N$-dimensional input:
\begin{align}
    \b{x} &\sim \mathcal{N}(0, \b{I}_N)\\
    \b{w}^* &\sim \mathcal{N}(0, \b{I}_N / \sqrt{N})\\
    y^*(\b{x}) &:= \text{sgn}({\b{w}^*}^T \b{x}).
\end{align}

The student is similarly a linearly network with parameters $\b{w}$, and it produces a policy through a sigmoidal nonlinearity:
\begin{align}
    \pi(y = 1; \b{x}) &= \sigma(\b{w}^T \b{x})\\
    \pi(y=-1; \b{x}) &= 1-\sigma(\b{w}^T \b{x})\\
    \sigma(z) &= \frac{1}{1 + e^{-z}}.
\end{align}
A single trial consists of a sequence of $T$ independent `inputs' and corresponding actions.
The agent receives a reward of $R = 1$ only if the entire sequence of choices is correct:
\begin{equation}
    R(y_{1:T} ; y^*_{1:T}) = \prod_t [y_t = y^*_t].
\end{equation}

We denote the pre-activations $z = \b{w}^T \b{x}$ and $z^* = {\b{w}^*}^T \b{x}$.
Our pre-activations are jointly Gaussian with
\begin{align}
    \sigma_{|*}^2. := \braket{z z^*} &= \braket{\b{w}^T \b{x} \b{x}^T \b{w}^*} \\
    &= \b{w}^T \braket{\b{x} \b{x}^T } \b{w}^*\\
    &= \b{w}^T \b{w}^*\\
    \sigma_{||}^2 := \braket{z z} &= \b{w}^T \b{w} .\\
    \sigma_{**}^2 := \braket{z^* z^*} &= {\b{w}^*}^T \b{w}^*.
\end{align}
The correlation between $z$ and $z^*$ is given by $\rho_{|*} := \frac{\sigma_{|*}^2}{\sigma_{**} \sigma_{||}}$.

If we sample actions greedily from the student, we can calculate the probability that a single action is correct, $p_g$, as a function of the student-teacher overlap $\rho_{| *}$.
For a zero-mean bivariate Gaussian distribution, the orthant formula tells us that
\begin{equation}
    p(z >0, z^* > 0) = \frac{1}{4} + \frac{\text{arcsin}(\rho)}{2 \pi}.
\end{equation}
This means the probability of the sstudent being correct is
\begin{align}
    p_g(\rho) &= p(\theta(z) = \theta(z^*))\\
    &= p(z >0, z^* > 0) + p(z < 0, z^* < 0)\\
    &= p(z >0, z^* > 0) + p(z >0, z^* > 0)\\
    &= \frac{1}{2} + \frac{\text{arcsin}(\rho_{|*})}{\pi}. \label{eq:pg}
\end{align}
Importantly, this is independent of any scaling of the weight vectors and depends only on their correlation.

We also remind ourselves that the gradient of the logistic function is given by
\begin{align}
    \frac{d \sigma (z)}{dz} &= \frac{e^{-z}}{(1 + e^{-z})^2}\\
    &= \left ( \frac{1}{1 + e^{-z}} \right ) \left ( \frac{e^{-z}}{1 + e^{-z}} \right ) \\
    &= \sigma(z) (1 - \sigma(z)).
\end{align}
Additionally,
\begin{align}
    \nabla_z \log \pi(y = 1) &= \nabla_z \log \sigma(z)\\
    &= (1 - \sigma (z))\\
    \nabla_z \log \pi (y = -1) &= \nabla_z \log (1 - \sigma(z))\\
    &= - \sigma (z)\\
    \nabla_z \log \pi (y = \tilde{y}) &= 0.5 + 0.5 \tilde{y} - \sigma(z).
\end{align}
At various times, we will approximate the sigmoid as a step function ($\sigma(z) \approx \theta (z)$) or linearise it ($\sigma(z) \approx 0.5 + 0.25 z$) for analytical tractability.

We also note a few integrals that will prove useful later.
For $x \sim \mathcal{N}(0, 1)$:
\begin{align}
    \braket{\theta (\b{v}^T \b{x}) \b{x} } &= \frac{1}{2} \braket{ \hat{\b{v}}^T \b{x} | \b{v}^T \b{x} > 0 } \hat{\b{v}}\\
    &= \frac{1}{2} \braket{\text{abs}(z)}_{z \sim \mathcal{N}(0, 1)} \hat{\b{v}}\\
    &= \sqrt{\frac{1}{2 \pi}} \hat{\b{v}}\\
    &:= \beta \hat{\b{v}}.
\end{align}
Since $\text{sgn}(z) = 2 \theta(z) - 1$, it follows that $\braket{\text{sgn} (\b{v}^T \b{x}) \b{x} } = 2 \beta \tilde{\b{v}}$.
We can also evaluate similar integrals conditional on agreement between the teacher and a greedy student, with the student and teacher vectors separated by an angle $\theta = \text{arccos}(\rho)$:
\begin{align}
    \label{eq:conditonal_sgn}
    \braket{\text{sgn}(\b{v}^T \b{x}) \b{x} | \text{sgn}(\b{v}^T \b{x})= \text{sgn}(\b{u}^T \b{x})} &= \frac{\hat{\b{v}} + \hat{\b{u}}}{|| \hat{\b{v}} + \hat{\b{u}} || } \frac{\int_{r = 0}^{\infty} \int_{\phi = - \frac{\pi - \theta}{2}}^{\frac{\pi - \theta}{2}} e^{-r^2/2} \; r \cos \phi  \; r dr \; d \phi}{\int_{r = 0}^{\infty} \int_{\phi = - \frac{\pi - \theta}{2}}^{\frac{\pi - \theta}{2}} e^{-r^2/2} r dr d \phi} \\ 
    &= \frac{\hat{\b{v}} + \hat{\b{u}}}{ \sqrt{ 1 + 1 + 2 \cos \theta } } \frac{\sqrt{\pi / 2} [\sin \phi]_{\phi = - \frac{\pi - \theta}{2}}^{\frac{\pi - \theta}{2}} }{ [e^{-r^2/2}]_{r=0}^\infty [\phi]_{\phi = - \frac{\pi - \theta}{2}}^{\frac{\pi - \theta}{2}} }\\
    &= \frac{\hat{\b{v}} + \hat{\b{u}}}{ \sqrt{2 + 2 \rho} } \sqrt{\pi / 2} \frac{2 \sin (\frac{\pi - \theta}{2}) }{\pi - \theta}\\
    &= \frac{\hat{\b{v}} + \hat{\b{u}}}{ \sqrt{ 1 + \rho} } \sqrt{ \pi} \frac{\sin (\frac{\pi - \theta}{2}) }{\pi - \theta}\\
    &:= \kappa ( \hat{\b{v}} + \hat{\b{u}} ).
\end{align}
The first line follows from noting that the $y = y^*$ condition is only satisfied for two cones with identical expectations, each of width $\pi - \theta$.
Within each cone, the vectors average out to the bisection, and the projection of a vector onto this line has magnitude $r \cos \phi$.
We integrate this projection over the area of interest.
We can similarly derive
\begin{align}
    \label{eq:conditonal_lin}
    \braket{(\hat{\b{v}}^T \b{x}) \b{x} | \text{sgn}(\b{v}^T \b{x})= \text{sgn}(\b{u}^T \b{x})} &= \hat{\b{v}} + \frac{\sin (\theta) }{\pi - \theta} \hat{\b{u}}\\
    &= \hat{\b{v}} + \frac{ \sqrt{1-\rho^2} }{\pi - \theta} \hat{\b{u}}.
\end{align}

\subsection*{Self-supervised learning}

We first consider the learning dynamics when doing supervised learning on training targets specified by an internal `planner' that takes the form of a simple sequence sampler and evaluator.
We will learn from a target action $\tilde{y}$ via the following loss function:
\begin{equation}
    \mathcal{L}(\b{w}) := \mathbb{E}_{ \{ \b{x}_t \sim p(\b{x}_t) \} } \left [ T^{-1} \sum_t \log \pi_t (y_t = \tilde{y}_t) \right ].
\end{equation}
Our gradient steps take the form
\begin{align}
    \nabla_{\b{w}} \mathcal{L}(\b{w}) &= T^{-1} \sum_t \nabla_{\b{w}} \log \pi_t (y_t = \tilde{y}_t) \\
    &= T^{-1} \sum_t (0.5-\sigma(z_t)+0.5 \tilde{y}_t) \b{x}_t.
\end{align}
For simplicity, we will approximate $\sigma(z) \approx \theta(z)$, where $\theta(z) = [z > 0]$ is the Heaviside step function.
Alternatively, we could have linearised $\theta (z)$ (see below).

Since all inputs and therefore actions are independent, we just consider the expected weight update when learning from a single action sampled from a sequence of $T$ action.
Given a probability $p_c$ of a target action being correct, we can write down the expected weight update
\begin{align}
    \braket{\Delta \b{w}} &= \braket{(0.5-\theta(z)+0.5 \tilde{y}) \b{x}} \\
    &= 0.5 \braket{\tilde{y} \b{x}} - \braket{\theta(z) \b{x}}\\
    &= 0.5 p_c \braket{y^* \b{x} | \tilde{y} = y^*} + 0.5 (1-p_c) \braket{ - y^* \b{x} | \tilde{y} \neq y^*} - \beta \hat{\b{w}}\\
    &\approx 0.5 p_c \braket{y^* \b{x}} + 0.5 (p_c-1) \braket{ y^* \b{x}} - \beta \hat{\b{w}}\\
    &= (2 p_c - 1) \beta \hat{\b{w}}^* - \beta \hat{\b{w}}.
\end{align}
Importantly, the penultimate line assumes that $p(\b{x} | \tilde{y}=y^*) = p(\b{x}) = \mathcal{N}(0, \b{I}_N)$.
For now, we will therefore assume that the planner simply samples correct actions with the same probability as a greedy student (\Cref{eq:pg}).
However, the actions themselves do not depend on the input $\b{x}$.

The overlap of the weight update with the goal vector is then
\begin{align}
    \braket{ (\Delta \b{w})^T \b{w}^*} &= \beta \left [  (2 p_c - 1) \sigma_{**}^{-1} {\b{w}^*}^T \b{w}^* - \sigma_{||}^{-1} \b{w}^T \b{w}^* \right ] \\ 
     &= \beta \left [  (2 p_c - 1) \sigma_{**} - \frac{\sigma_{|*}^2}{\sigma_{||}} \right ].
\end{align}
The overlap with the student vector is
\begin{align}
    \braket{ (\Delta \b{w})^T \b{w}} &= \beta \left [  (2 p_c - 1) \sigma_{**}^{-1} \b{w}^T \b{w}^* - \sigma_{||}^{-1} \b{w}^T \b{w} \right ] \\ 
     &= \beta \left [  (2 p_c - 1) \frac{\sigma_{|*}^2}{\sigma_{**}} - \sigma_{||} \right ].
\end{align}
The overlap with the difference is
\begin{align}
    \braket{ (\b{w}^* - \b{w})^T (\Delta \b{w})} &= \beta \left [  (2 p_c - 1) \sigma_{**} + \sigma_{||} - \frac{\sigma_{|*}^2}{\sigma_{||}} - (2 p_c - 1) \frac{\sigma_{|*}^2}{\sigma_{**}} \right ].
\end{align}

We now compute the expected change in the different $\sigma$s when updating $\b{w} \leftarrow \b{w} + \eta \Delta \b{w}$.
\begin{align}
    \braket{\Delta \sigma_{**}^2} &= \braket{\Delta ({\b{w}^*}^T \b{w}^*)}\\
    &= 0,
\end{align}
since $\b{w}^*$ is fixed.
\begin{align}
    \braket{\Delta \sigma_{|*}^2} &= \braket{\Delta (\b{w}^T \b{w}^*)}\\
    &= \braket{(\b{w} + \eta \Delta \b{w})^T \b{w}^* - \b{w}^T \b{w}^*}\\
    &= \eta \braket{(\Delta \b{w})^T \b{w}^*}\\
    &= \eta \beta \left [  (2 p_c - 1) \sigma_{**} - \frac{\sigma_{|*}^2}{\sigma_{||}} \right ].
\end{align}
\begin{align}
    \braket{\Delta \sigma_{||}^2} &= \braket{\Delta (\b{w}^T \b{w})}\\
    &= \braket{(\b{w} + \eta \Delta \b{w})^T (\b{w} + \eta \Delta \b{w}) - \b{w}^T \b{w}}\\
    &= \braket{2 \eta (\Delta \b{w})^T \b{w} + \eta^2 (\Delta \b{w})^T (\Delta \b{w})}\\
    &\approx 2 \eta \braket{(\Delta \b{w})^T \b{w}}\\
    &= 2 \eta \beta \left [  (2 p_c - 1) \frac{\sigma_{|*}^2}{\sigma_{**}} - \sigma_{||} \right ],
\end{align}
to first order in the learning rate $\eta$.

To write down the full learning dynamics, we also need to compute $p_c$.
Recall that every simulated action is sampled independently, and that it is correct with probability $p_g$.
The probability that at least one sampled sequence is entirely correct is given by
\begin{equation}
    p_R := p( \exists_{i \in [1, 2, \ldots, M]} \; \text{s.t.} \; \tilde{y}_{it} = {y_t^*} \; \forall_t) = 1 - (1 - p_g^T)^M.
\end{equation}
In this case, all of the individual actions are correct.
The probability that no sampled sequences are entirely correct is $(1 - p_g^T)^M$.
In this case a random sequence sampled from the set of `not entirely correct' sequences is used as a target.
The marginal probability of an action being correct, given that the entire sequence is not, is:
\begin{align}
    p_0^* := p(y_t = y^* | \text{not all correct}) &= \frac{p(\text{not all correct}| y_t = y^*) p(y_t = y^*)}{p(\text{not all correct})}\\
    &= \frac{(1 - p_g^{T-1}) p_g}{1 - p_g^T}
\end{align}
We therefore get
\begin{equation}
    \label{eq:pc}
    p_c = 1 - (1 - p_g^T)^M + (1 - p_g^T)^M \frac{(1 - p_g^{T-1}) p_g}{1 - p_g^T}.
\end{equation}

Finally, we can write down the full set of learning dynamics, where we set $\sigma_{**} = 1$ without loss of generality:
\begin{align}
    \rho_{|*} &= \frac{\sigma_{|*}^2}{\sigma_{||}}\\
    p_g &= \frac{1}{2} + \frac{\text{arcsin}(\rho_{|*})}{\pi}\\
    p_R &= 1 - (1 - p_g^T)^M\\
    p_c &= p_R + (1-p_R) \frac{(1 - p_g^{T-1}) p_g}{1 - p_g^T}\\
    \Delta \sigma_{|*}^2 &= \eta \beta \left [  (2 p_c - 1) - \frac{\sigma_{|*}^2}{\sigma_{||}} \right ]\\
    \Delta \sigma_{||}^2 &= 2 \eta \beta \left [  (2 p_c - 1) \sigma_{|*}^2 - \sigma_{||} \right ].
\end{align}

\subsubsection*{Supervised learning}

We recover supervised learning simply by letting $M \rightarrow \infty$ so $p_c \rightarrow 1$.
In this case, we have
\begin{align}
    p_R &= p_g^T\\
    \Delta \sigma_{|*}^2 &= \eta \beta \left [  1 - \frac{\sigma_{|*}^2}{\sigma_{||}} \right ]\\
    \Delta \sigma_{||}^2 &= 2 \eta \beta \left [  \sigma_{|*}^2 - \sigma_{||} \right ].
\end{align}
Here, $p_R$ still denotes the probability of the agent generating a correct action sequence and getting reward.
It comes from noting that the supervised learner only produces one sequence in a trial, and every element has to be correct for the sequence to be correct.

\subsubsection*{Reinforcement learning}

We now consider a reinforcement learner.
The policy gradient is given by:
\begin{equation}
    \nabla_{\b{w}} \mathcal{L}(\b{w}) = \mathbb{E}_{ \{ \b{x}_t \sim p(\b{x}_t) \} } \left [ T^{-1} \sum_t (R - B) \nabla_{\b{w}} \log \pi_t (y = y_t) \right ],
\end{equation}
where $B$ is a baseline.
In this case, the `scale' of $z$ matters.
The policy is approximately linear when $z$ has small variance, and it approaches a step function when $z$ has large variance.
For now, we will use the `normalised' $\hat{z} = \sigma_{||}^{-1} z$.
A single gradient step thus takes the form
\begin{align}
    \Delta \b{w} &= (R - B) (0.5-\sigma(\hat{z})+0.5 y) \b{x},\\
    \Delta \b{w} &\approx (R - B) (0.5 y - 0.25 \hat{z}) \b{x}.
\end{align}
where we have linearised $\sigma(\hat{z}) \approx 0.5 + 0.25 \hat{z}$.

We now compute the expected weight update
\begin{align}
    \braket{\Delta \b{w}} &= \braket{(R - B) (0.5 y - 0.25 \hat{z}) \b{x}}\\
    &= 0.5 \braket{R y \b{x}} - 0.25 \braket{R \hat{z} \b{x}} + 0.25 B \braket{\hat{z} \b{x}} - 0.5 B \braket{y \b{x}}\\
    &= 0.5 p_R \braket{y \b{x} | y = y^*} - 0.25 p_R \braket{ \hat{z} \b{x} | y = y^*} + 0.25 B \braket{\hat{z} \b{x}} - 0.5 B \braket{y \b{x}}.
\end{align}
To evaluate the conditional expectations, we will assume that the agent takes greedy actions, $y = \text{sgn}(z)$ (although we still assume a linearised sigmoid in the gradient, which is a bit weird).
Using \Cref{eq:conditonal_sgn} and \Cref{eq:conditonal_lin}, we get
\begin{align}
    \braket{\Delta \b{w}} &= 0.5 \kappa p_R (\hat{\b{w}} + \hat{\b{w}}^*) - 0.25 p_R \left (\hat{\b{w}} + \frac{\sqrt{1-\rho^2} }{\pi - \theta} \hat{\b{w}}^* \right ) + 0.25 B \hat{\b{w}} - B \beta \hat{\b{w}}\\
    &= \frac{\b{w}}{\sigma_{||}} \left ( 0.5 p_R \kappa - 0.25 p_R  + 0.25 B - \beta B \right ) + \frac{\b{w}^*}{\sigma_{**}} \left ( 0.5 p_R \kappa - 0.25 p_R \frac{\sqrt{1-\rho^2} }{\pi - \theta} \right ).
\end{align}
The overlap with the student and target vectors are:
\begin{align}
    \braket{ (\Delta \b{w})^T \b{w}^*} &= \frac{\sigma_{|*}^2}{\sigma_{||}} \left ( 0.5 p_R \kappa - 0.25 p_R  + 0.25 B - \beta B \right ) + \sigma_{**} \left ( 0.5 p_R \kappa - 0.25 p_R \frac{\sin (\theta) }{\pi - \theta}  \right ),\\
    \braket{ (\Delta \b{w})^T \b{w}} &= \sigma_{||} \left ( 0.5 p_R \kappa - 0.25 p_R  + 0.25 B - \beta B \right ) + \frac{\sigma_{|*}^2}{\sigma_{**}} \left ( 0.5 p_R \kappa - 0.25 p_R \frac{\sqrt{1-\rho^2} }{\pi - \theta} \right )
\end{align}
The full learning dynamics with $\sigma_{**} = 1$ are:
\begin{align}
    \rho_{|*} &= \frac{\sigma_{|*}^2}{\sigma_{||}}\\
    p_g &= \frac{1}{2} + \frac{\text{arcsin}(\rho_{|*})}{\pi}\\
    p_R &= p_g^T \\
    \kappa &= \frac{ \sqrt{\pi} }{ \sqrt{ 1 + \rho} } \frac{\sin (\frac{\pi - \theta}{2}) }{ (\pi - \theta) }\\
    \Delta \sigma_{|*}^2 &= \eta \left [ \frac{\sigma_{|*}^2}{\sigma_{||}} \left ( 0.5 p_R \kappa - 0.25 p_R  + 0.25 B - \beta B \right ) + \left ( 0.5 p_R \kappa - 0.25 p_R \frac{\sqrt{1-\rho^2} }{\pi - \theta}  \right ) \right ] \\
    \Delta \sigma_{||}^2 &= 2 \eta \left [ \sigma_{||} \left ( 0.5 p_R \kappa - 0.25 p_R  + 0.25 B - \beta B \right ) + \sigma_{|*}^2 \left ( 0.5 p_R \kappa - 0.25 p_R \frac{ \sqrt{1-\rho^2} }{\pi - \theta} \right ) \right ].
\end{align}

\subsubsection*{Correlation changes}

We can also compute the change in correlation as a function of the current correlation for a normalised student+teacher:
\begin{align}
    \dot{\rho} &= \frac{\rho + \dot{\sigma_{|*}}}{\sqrt{1+ \dot{\sigma_{||}}}} - \rho\\
    &\approx (\rho + \dot{\sigma_{|*}})(1-0.5 \dot{\sigma_{||}}) - \rho\\
    &\approx \dot{\sigma_{|*}} - 0.5 \dot{\sigma_{||}},
\end{align}
to first order in $\eta$.
We can then write
\begin{align}
    \dot{\rho}_\text{sup} &= \beta ( 1 - \rho^2  )\\
    \dot{\rho}_\text{pred} &= \beta (2p_c-1)(1  - \rho^2)\\
    \dot{\rho}_\text{RL} &= \left ( 0.5 p_g^T \kappa - 0.25 p_g^T \frac{\sqrt{1-\rho^2}}{\pi- \theta} \right ) (1 - \rho^2 ).
\end{align}

It is reasonably straightforward to both (i) linearise these expressions, and (ii) calculate their maximum.
If we linearise, we get an exponential curve that explains early learning.


\subsubsection*{Gradient signal and noise}

Now let's compute the variance of our gradient estimators.
We will assume that $N \rightarrow \infty$, so $x_{it}$ is uncorrelated from $z$ and $y$, which neglects terms in $N^{-1}$.

\begin{align}
    \braket{\Delta w_i^2}_{x} &= \braket{((0.5 - \theta (z) + 0.5 \tilde{y})x_i)^2}\\
    &= 0.25 \braket{(\tilde{y} - y)x_{i}^2}\\
    &\approx 0.25 (\braket{\tilde{y}^2} + \braket{y^2} - 2 \braket{\tilde{y} y})\\
    &= 0.5 - 0.5 (p_c p_g + (1-p_c)(1-p_g) - p_g(1-p_c) - p_c(1-p_g))\\
    &= 0.5 - 0.5 (2 p_g - 1) (2 p_c - 1).
\end{align}
This holds for \emph{all} elements of the weight vector.
In the supervised learning case $p_c = 1.0$ and this reduces to $1 - p_g$.
At initialisation, $p_g = 0.5$ and we are left with $0.5$.
Importantly, the only effect of $T$ in the predictive learning case is to reduce the magnitude of the second term since $p_c \rightarrow 0.5$.
However, the second moment is still capped at 0.5.
The expected square of the mean updates is
\begin{align}
    \braket{\braket{\Delta w_i}_x^2}_i &= \braket{((2 p_c - 1) \beta \hat{w}^*_i - \beta \hat{w}_i)^2}_i\\
    &= N^(-1) \beta^2 ( 1 + (2p_c-1)^2 - 2 (2 p_c - 1) \rho ).
\end{align}
The $N^{-1}$ term arises because each element of our student and teacher scales with $N^{-\frac12}$ to ensure normalisation.
The scaled noise-to-signal ratio is
\begin{align}
    \gamma = N^{-1} \frac{\braket{\Delta w_i^2}_{x} - \braket{\braket{\Delta w_i}_x^2}_i}{\braket{\braket{\Delta w_i}_x^2}_i}\\
    &\approx \frac{0.5 - 0.5 (2 p_g - 1) (2 p_c - 1)}{\beta^2 ( 1 + (2p_c-1)^2 - 2 (2 p_c - 1) \rho )}.
\end{align}
Notably, this is approximately constant in $T$.

In the RL case
\begin{align}
    \braket{\Delta w_i} &= \braket{((R-B)( 0.5 y - 0.25 \hat{z}) x_i)^2}\\
    &\approx \braket{(R-B)^2} (\frac{1}{4} \braket{y^2} + \frac{1}{16} \braket{\hat{z}^2} - \frac{1}{4} \braket{\text{sgn}(\hat{z} \hat{z})})\\
    &= (p_R + B^2 - 2Bp_R)(\frac{5}{16} + \frac{1}{2}\beta)\\
    &\geq 0.11 p_R(1-p_R).
\end{align}
Conversely, 
\begin{align}
    \braket{\braket{\Delta w_i}_x^2}_i &= \braket{ \left ( \hat{\b{w}}_i \left ( 0.5 p_R \kappa - 0.25 p_R  + 0.25 B - \beta B \right ) + \hat{\b{w}^*}_i \left ( 0.5 p_R \kappa - 0.25 p_R \frac{\sqrt{1-\rho^2} }{\pi - \theta} \right ) \right )^2 }\\
    &= \left ( 0.5 p_R \kappa - \beta p_R \right )^2 + \left ( 0.5 p_R \kappa - 0.25 p_R \frac{\sqrt{1-\rho^2} }{\pi - \theta} \right )^2 + \rho \left ( 0.5 p_R \kappa - \beta p_R \right ) \left ( 0.5 p_R \kappa - 0.25 p_R \frac{\sqrt{1-\rho^2} }{\pi - \theta} \right )\\
    &= p_R^2 \left [ \left ( 0.5 \kappa - \beta \right )^2 + \left ( 0.5 \kappa - 0.25 \frac{\sqrt{1-\rho^2} }{\pi - \theta} \right )^2 + \rho \left ( 0.5 \kappa - \beta \right ) \left ( 0.5 \kappa - 0.25 \frac{\sqrt{1-\rho^2} }{\pi - \theta} \right ) \right ].
\end{align}
Now the scaled signal-to-noise ratio is
The scaled noise-to-signal ratio is
\begin{align}
    \gamma = N^{-1} \frac{\braket{\Delta w_i^2}_{x} - \braket{\braket{\Delta w_i}_x^2}_i}{\braket{\braket{\Delta w_i}_x^2}_i}\\
    &\approx p_R^{-1} \frac{0.11 (1-p_R)}{\left ( 0.5 \kappa - \beta \right )^2 + \left ( 0.5 \kappa - 0.25 \frac{\sqrt{1-\rho^2} }{\pi - \theta} \right )^2 + \rho \left ( 0.5 \kappa - \beta \right ) \left ( 0.5 \kappa - 0.25 \frac{\sqrt{1-\rho^2} }{\pi - \theta} \right )}.
\end{align}
Notably, this scales as $p_R = p_g^T$!
We therefore see that both supervised and predictive learning have a constant signal-to-noise ratio, while RL is exponentially bad.

\newpage

\subsubsection*{Linearised gradient}

It is also straightforward to derive the `predictive' learning dynamics with $\sigma (z) \approx 0.5 + 0.25 z$.
\begin{align}
    \braket{ \Delta \b{w}} &= \braket{(0.5 \tilde{y} -\sigma(\hat{z}) + 0.5) \b{x}}\\
    &\approx 0.5 \braket{\tilde{y}} - 0.25 \braket{\hat{z} \b{x}}\\
    &= p_c \beta \hat{\b{w}}^* - (1-p_c)\beta \hat{\b{w}}^* - 0.25 \hat{\b{w}}\\
    &= (2 p_c - 1) \beta \hat{\b{w}}^* - 0.25 \hat{\b{w}}.
\end{align}

The learning dynamics are then:
\begin{align}
    \Delta \sigma_{|*}^2 &= \eta \left [  \beta (2 p_c - 1) - 0.25 \frac{\sigma_{|*}^2}{\sigma_{||}} \right ]\\
    \Delta \sigma_{||}^2 &= 2 \eta \left [  \beta (2 p_c - 1) \sigma_{|*}^2 - 0.25 \sigma_{||} \right ].
\end{align}

With supervised learning, this reduces to
\begin{align}
    \Delta \sigma_{|*}^2 &= \eta \left [  \beta - 0.25 \frac{\sigma_{|*}^2}{\sigma_{||}} \right ]\\
    \Delta \sigma_{||}^2 &= 2 \eta \left [  \beta \sigma_{|*}^2 - 0.25 \sigma_{||} \right ].
\end{align}

\subsubsection*{Miscellaneous}

When linearising, we could also minimise the expected error
\begin{align}
    \mathcal{L} &= \mathbb{E} [(\sigma(z) - (az + 0.5))^2]\\
    &= \mathbb{E} [(\sigma(z) - az - 0.5)^2]
\end{align}
empirically, this gives $a = 0.207$.

\newpage

% Our reinforcement learning update takes the form
% \begin{equation}
%     \nabla_{\b{w}} \mathcal{L}(\b{w}) = \mathbb{E}_{ \{ \b{x}_t \sim p(\b{x}_t) \} } \left [ T^{-1} \sum_t (R - B) \nabla_{\b{w}} \log \pi_t (y = y_t) \right ],
% \end{equation}
% where $B$ is a baseline.
% The variance of the parameter update is minimised when the baseline is the expected reward, $B = p_R$, and we will use this for simplicity.
% As in the supervised case, the probability of producing a correct sequence is simply
% \begin{equation}
%     p_R = p_g^T.
% \end{equation}
% A single gradient step thus takes the form
% \begin{align}
%     \Delta \b{w} &= (R - p_R) (0.5-\sigma(z)+0.5 y) \b{x},\\
%     \Delta \b{w} &\approx (R - p_R) (0.5 y - 0.25 z) \b{x}.
% \end{align}
% where we have again set $\sigma(z) \approx \theta(z)$.

% We will consider three different cases;
% \begin{align}
%     \Delta \b{w} | R = 1, y = y^* \ &= (1 - p_R) (0.5 y^* - 0.25 z) \b{x},\\
%     \Delta \b{w} | R = 0, y = y^* \ &= - p_R (0.5 y^* - 0.25 z) \b{x},\\
%     \Delta \b{w} | R = 0, y \neq y^* \ &= - p_R (- 0.5 y^* - 0.25 z) \b{x}.
% \end{align}
% These happen with probabilities
% \begin{align}
%     p(R=1, y = y^*) &= p_R \\
%     p(R = 0, y = y^*) &= (1-p_R) p_0^*\\
%     p(R = 0, y \neq y^*) &= (1-p_R) (1 - p_0^*),
% \end{align}
% where $p_0^* = p(y_t = y^* | \text{not all correct}) = \frac{(1 - p_g^{T-1}) p_g}{1 - p_g^T}$ as before.

% To compute the expected weight update, we note that $\theta = \text{cos}^{-1} (\rho)$ and
% \begin{align}
%     \mathbb{E} \left [ z | y = y^* \right ] &= \mathbb{E} \left [ \b{w}^T \b{x} | \text{sgn}(\b{w}^T \b{x}) = \text{sgn}( {\b{w}^*}^T \b{x}) \right ] \\
%     &= \b{w} + \frac{\sin (\theta) }{\pi - \theta} \frac{\sigma_{||}}{\sigma_{**}} \b{w}^*\\
%     &= \b{w} + \kappa_1 \b{w}^*.
% \end{align}
% Conversely,
% \begin{align}
%     \mathbb{E} \left [ z | y \neq y^* \right ] &= \b{w} - \frac{\sin (\theta) }{ \theta} \frac{\sigma_{||}}{\sigma_{**}} \b{w}^*\\
%     &= \b{w} + \kappa_2 \b{w}^*.
% \end{align}
% Additionally, 
% \begin{align}
%     \mathbb{E} \left [ \text{sgn}(y^*) | y = y^* \right ] &= \sqrt{\frac{2}{\pi}} \frac{1}{2 \cos(\theta/2)} (\hat{\bf{w}} + \hat{\bf{w^*}})\\
%     &= kappa_3 (\hat{\bf{w}} + \hat{\bf{w^*}})
% \end{align}
% and
% \begin{align}
%     \mathbb{E} \left [ \text{sgn}(y^*) | y = y^* \right ] &= \sqrt{\frac{2}{\pi}} \frac{1}{2 \sin(\theta/2)} (\hat{\bf{w}} - \hat{\bf{w^*}})\\
%     &= \kappa_4 (\hat{\bf{w}} - \hat{\bf{w^*}}).
% \end{align}
% This allows us to write down the expected weight update
% \begin{align}
%     \braket{\Delta \b{w}} &= p_R (1 - pR) (0.5 \kappa_3  - 0.25 \kappa_1) - p_R 
%     (1-p_R) p_0^* (0.5 \kappa_3  - 0.25 \kappa_1) - pR (1-p_R) (1 - p_0^*) (- 0.5 \kappa_4  - 0.25 \kappa_2)
% \end{align}


% This allows us to write down the expected weight update
% \begin{align}
%     \braket{\Delta \b{w}} &= \braket{ ( p_R (0.5-\theta(z)) - p_R (0.5 - \theta (z))) \b{x}} \\
%     &+ \braket{0.5 y^* (p_R(1-p_R) + (1-p_R)p_0^* (-p_R) + (1-p_R)(1-p_0^*)(-p_R)(-1)) \b{x} }\\
%     &= \beta \sigma_{**}^{-1} \b{w}^* (p_R-p_R^2 - p_0^*p_R + p_R^2 p_0^* + p_R + p_R^2 p_0^* - p_R^2 - p_R p_0^*)\\
%     &= \beta \sigma_{**}^{-1} \b{w}^* (2 p_R - 2 p_R^2 - 2 p_R p_0^* + 2 p_R^2 p_0)\\
%     &= 2 \beta p_R \sigma_{**}^{-1} \b{w}^* (1 + p_R p_0^* - p_R - p_0^*).
% \end{align}
% the first line comes from noting that (i) the probabilities of the three options sum to 1, and they each have a term of the form $-p_R (0.5 - \theta (z))$; and (ii) the first case contributes an update $(0.5 - \theta (z))$ with probability $p_R$.

% The overlap with the student and target vectors are:
% \begin{align}
%     \braket{ (\Delta \b{w})^T \b{w}^*} &= 2 \beta p_R (1 + p_R p_0^* - p_R - p_0^*) \sigma_{**},\\
%     \braket{ (\Delta \b{w})^T \b{w}} &= 2 \beta p_R (1 + p_R p_0^* - p_R - p_0^*) \frac{\sigma_{|*}^2}{\sigma_{**}}.
% \end{align}
% The full learning dynamics with $\sigma_{**} = 1$ are:
% \begin{align}
%     \rho_{|*} &= \frac{\sigma_{|*}^2}{\sigma_{||}}\\
%     p_g &= \frac{1}{2} + \frac{\text{arcsin}(\rho_{|*})}{\pi}\\
%     p_R &= p_g^T \\
%     p_0^* &= \frac{(1 - p_g^{T-1}) p_g}{1 - p_g^T}\\
%     \Delta \sigma_{|*}^2 &= 2 \eta \beta p_R (1 + p_R p_0^* - p_R - p_0^*) \\
%     \Delta \sigma_{||}^2 &= 4 \eta \beta p_R (1 + p_R p_0^* - p_R - p_0^*) \sigma_{|*}^2.
% \end{align}


% If we set the baseline to zero instead, we would have
% \begin{align}
%     p_R &= p_g^T\\
%     \braket{\Delta w} &= \beta p_R (\sigma_{**} \b{w}^* - \sigma_{||} \b{w}) \\
%     \Delta \sigma_{|*}^2 &= \eta \beta p_R \left [  1 - \frac{\sigma_{|*}^2}{\sigma_{||}} \right ]\\
%     \Delta \sigma_{||}^2 &= 2 \eta \beta p_R \left [  \sigma_{|*}^2 - \sigma_{||} \right ].
% \end{align}

\end{document}